Explore the SingleR results

### 2025-07-15 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd
add final section boxplot

Also, note that the `myeloid`-labeled SingleR cells are _only_ neutrophils, but the consensus myeloid cagetory contains additonal cell types.

@@

@@

situation_df <- celltype_recoded_df |>
  # recode labels for comparison
  dplyr::mutate(
    label_recoded = ifelse(
      label_recoded %in% consensus_myeloid,
      "myeloid",
      label_recoded
  )) |>
  dplyr::filter(label_recoded %in% direct_celltype_matches) |>
  dplyr::select(cell_id, annotation_type, label_recoded) |>
  # make it wider to label the situation
  tidyr::pivot_wider(
    names_from = "annotation_type",
    values_from = "label_recoded"
  ) |>
  # remove NAs:
  # - consensus NAs are those with other cell types not to include here
  # - singler NAs were unclassified cells
  tidyr::drop_na() |>
  dplyr::mutate(
    annotation_situation = dplyr::case_when(
      consensus_validation == "unknown_consensus" ~ "consensus not classified", 
      consensus_validation == singler ~ "matching annotations",
      consensus_validation != singler ~ "different annotations"
    )
  ) |>
  # make it long again
  tidyr::pivot_longer(
    c(consensus_validation, singler),
    names_to = "annotation_type", 
    values_to = "label_recoded"
  )

cells <- unique(situation_df$cell_id)
genes <- unique(consensus_markers_df$ensembl_gene_id)
```

```{r}
scuttle::makePerCellDF(
    merged_sce[genes, cells],
    features = genes,
    assay.type = "logcounts",
    use.coldata = "cell_id",
    use.dimred = FALSE
  ) |>
    tidyr::pivot_longer(starts_with("ENSG"), names_to = "ensembl_gene_id", values_to = "logcounts") |>
  dplyr::left_join(situation_df, by = "cell_id") |>
  dplyr::group_by(label_recoded, marker_gene_label, ensembl_gene_id) |>
  dplyr::summarize(
    detected_count = sum(detected),
    mean_exp = mean(logcounts)
  ) |>
  dplyr::ungroup()

ggplot(situation_df) + 
  aes(x = singler, y = )

  group_stats_df <- all_info_df |>
    # remove genes that aren't present in final annotations
    dplyr::filter(gene_symbol %in% marker_gene_order) |>
    # for each assigned cell type/marker gene combo get total detected and mean expression
    dplyr::group_by(label_recoded, marker_gene_label, ensembl_gene_id) |>
    dplyr::summarize(
      detected_count = sum(detected),
      mean_exp = mean(logcounts)
    ) |>

>>>>

Also, note that the `myeloid`-labeled SingleR cells are _only_ neutrophils, but the consensus myeloid grouping contains additional cell types.

@@

## Compare expression when SingleR and consensus annotations match

Next, we look at expression of consensus validation marker genes for each normal/harmonized cell type label.
We compare expression across groups of cells:

* `matching annotations`: Cells whose consensus annotation matches the SingleR annotation
* `different annotations`: Cells whose consensus annotation differs from the SingleR annotation
* `consensus not classified`: Cells where the consensus annotation is unknown

Each panel contains only cells that SingleR annotated with the given label, and shows only marker genes associated with that given cell type. 
We expect that cells whose consensus and SingleR annotations agree may show higher gene expression for the given marker.

@@

# Data frame with column `compare_annotations` that indicates
# if consensus and singler (roughly) match
compare_annotations_df <- celltype_recoded_df |>
  # recode labels for comparison
  dplyr::mutate(
    label_recoded = ifelse(
      label_recoded %in% consensus_myeloid,
      "myeloid",
      label_recoded
  )) |>
  dplyr::filter(label_recoded %in% direct_celltype_matches) |>
  dplyr::select(cell_id, annotation_type, label_recoded) |>
  # make it wider to label the situation
  tidyr::pivot_wider(
    names_from = "annotation_type",
    values_from = "label_recoded"
  ) |>
  # remove NAs:
  # - consensus NAs are those with other cell types not to include here
  # - singler NAs were unclassified cells
  tidyr::drop_na() |>
  dplyr::mutate(
    compare_annotations = dplyr::case_when(
      consensus_validation == "unknown_consensus" ~ "consensus not classified", 
      consensus_validation == singler ~ "matching annotations",
      consensus_validation != singler ~ "different annotations"
    ), 
    compare_annotations = forcats::fct_relevel(compare_annotations, "matching annotations", "different annotations", "consensus not classified")
  ) |>
  # we can remove the consensus annotation now
  dplyr::select(cell_id, singler, compare_annotations)

direct_markers_df <- consensus_markers_df |>
  dplyr::filter(validation_group_annotation %in% direct_celltype_matches) |>
  dplyr::select(marker_gene_label = validation_group_annotation, ensembl_gene_id, gene_symbol)
cells <- unique(compare_annotations_df$cell_id)
genes <- unique(direct_markers_df$ensembl_gene_id)
```

```{r}
compare_annotations_expr_df <- scuttle::makePerCellDF(
  merged_sce[genes, cells],
  features = genes,
  assay.type = "logcounts",
  use.coldata = "cell_id",
  use.dimred = FALSE
) |>
  tidyr::pivot_longer(starts_with("ENSG"), names_to = "ensembl_gene_id", values_to = "logcounts") |>
  # filter to only cells with expression
  dplyr::filter(logcounts > 0) |>
  # join other data frames
  dplyr::left_join(compare_annotations_df, by = "cell_id") |>
  dplyr::left_join(direct_markers_df, by = "ensembl_gene_id", relationship = "many-to-many")
```

```{r}
unique(compare_annotations_expr_df$singler) |>
  purrr::map(
    \(celltype) {

      plot_df <- compare_annotations_expr_df |>
        dplyr::filter(singler == celltype, marker_gene_label == celltype)

      ggplot(plot_df) + 
        aes(x = gene_symbol, y = logcounts, fill = compare_annotations) +
        geom_boxplot(outlier.size = 0.25) + 
        ggtitle(glue::glue("cells annotated as: {celltype}"))

  }) |>
  patchwork::wrap_plots(ncol = 1) + patchwork::plot_layout(guides = "collect")
```
